Generic Properties of D-optimal Designs in Infinite Dimensional Bayesian Linear Inverse Problems

### 2023-04-02 latex/vanishing.tex
Fixin
getting closegit add latex/vanishing.tex !

%% \subsection{The Nonlinear Eigenvalue Problem}\label{subsec:eigenproblem}
$\modcov = 0$ implies $\Sigma= \sigma^2I$. The necessary first-order
condition for D-optimality \eqref{eq:conditions} becomes
\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi,
\end{equation}
with $\Xi$ diagonal. We think of \eqref{eq:eigenproblem} as an
eigenvalue problem for the self-adjoint operator $\sigma^{-2}\fwd
\postcov \fwd^*$, where rows of $\obs$, namely $\meas_j,j=1,\dots, m$,
are eigenvectors. However, that $\postcov$ depends on $\obs$ and thus,
we refer to \eqref{eq:eigenproblem} as a nonlinear eigenvalue problem.

The following two lemmas allow us to gain better understanding of the
nonlinear eigenvalue problem.

@@

\begin{restatable*}[Simultaneous diagonizability for the nonlinear eigenvalue problem]{lemma}{simdiag}\label{lemma:sim diag}
  Let $\hil$ separable Hilbert space, $C:\hil \to \hil$ self-adjoint
  and $\func_1,\dots,\func_m \in \hil$. Denote $\func^*$ the element
  $\func$ acting as a linear functional. If
  \begin{equation*}
   (C + \sum_{j=1}^m \func_j\func_j^*) \func_l = \xi_l \func_l,\ l = 1,\dots,m
  \end{equation*}
  then $C$ and $\sum_{j=1}^m \func_j \func_j^*$ are simultaneously
  diagonalizable.
\end{restatable*}

\begin{corollary}[Eigenvectors of a D-optimal design]\label{cor:same ev}
  Let $\obs$ satisfy the nonlinear eigenvalue problem
  \eqref{eq:eigenproblem}. Then $\obs^*\obs$ has the same eigenvectors
  as $\fwd \prcov \fwd^*$.
\end{corollary}
\begin{proof}
  \begin{align}\label{eq:mod conditions}
    \begin{split}
      \obs^* \Xi &= \sigma^{-2}\fwd \postcov \fwd^* \obs^*  \text{ (by \eqref{eq:eigenproblem})}\\
      %
      %
      %
      &= \sigma^{-2} \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* \obs^*  \text{ (by \eqref{eq:postcov})} \\
      %
      %
      %
      &= \sigma^{-2} \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1} \obs^* \text{ (by Lemma \ref{lemma:twice woodbury})}.
    \end{split}
  \end{align}

@@

Corollary \ref{cor:same ev} implies we only need to consider $\obs$
such that at most $m$ of the eigenvectors of $\obs^*\obs$ have a
positive eigenvalue, because $k := \rank \obs^*\obs \leq m$. Since we
made no assumption regarding the ordering of $\{\lambda_i\}$, we can
denote the corresponding non-zero eigenvalues of $\obs^*\obs$ by
$\{\eta_i\}_{i=1}^{k}$ and let $\eta_i = 0$ for $i \geq k+1$.

%% \begin{definition}
%%   Let $\{\lambda_i\}_{i=1}^{\infty}$ eigenvalues of $\fwd \prcov
%%   \fwd^*$ sorted in descending order. Let $\eta \in
%%   \mathbb{R}_+^m$. The \emph{utility proxy} $\bar{\tar}$ is defined
%%   as:
%%   \begin{equation*}
%%     \bar{\tar}(\eta) := \frac12 \sum_{i=1}^m \log(1 + \sigma^{-2}\lambda_i\eta_i).
%%   \end{equation*}
%% \end{definition}

\begin{corollary}\label{cor:true target}
  Let $\obs$ with $m$ observations satisfy the nonlinear eigenvalue
  problem \eqref{eq:eigenproblem}. Let $\{\eta_i\}_{i=1}^{\infty}$
  eigenvalues of $\obs^*\obs$ and $\{\lambda_i\}_{i=1}^{\infty}$ the
  corresponding eigenvalues of $\fwd \prcov \fwd^*$. Let $k:=\rank
  \obs^*\obs$. Without loss of generality, let $\eta_i > 0$ for $i\leq
  k$ and $\eta_i = 0$ for $i > k$. Then:
  \begin{enumerate}
    \item $k \leq m$ and $\obs^*\obs$ has exactly $k$ positive
      eigenvalues.
    \item
      \begin{equation*}
        \tar(\obs) = \frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i) = \frac12 \sum_{i=1}^{m} \log (1 + \sigma^{-2}\lambda_i\eta_i).
      \end{equation*}
    \item Let $\obs$ a D-optimal design. Then $\eta_i > 0$ for
      eigenvectors corresponding to the $k$ largest $\lambda_i$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Part (1) is trivial. To see part (2) holds: 
  \begin{align}
    \begin{split}
      \tar(\obs) &= \frac12\log \det \left (I + \sigma^{-2} \prcov^{1/2} \fwd ^* \obs^*
      \obs \fwd \prcov^{1/2}\right )\\% \text{ (by definition)}\\
      %
      &= \frac12 \log \det \left (I + \sigma^{-2} \obs^* \obs \fwd
      \prcov\fwd^* \right ) \text{ (Sylvester's Determinant
      Theorem)}\\
      %
      %
      %
      &=\frac12 \log \prod_{i=1}^{\infty} ( 1 + \sigma^{-2} \lambda_i\eta_i ) \text{ (Corollary \ref{cor:same ev})} \\
      %
      %
      %
      %&=\frac12 \log \left ( \prod_{i=1}^{k} ( \lambda_i^{-1} + \sigma^{-2} \eta_i )\prod_{i=1}^{k} \lambda_i \right )
      %
      %
      %
      &=\frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i). 
      %
      %
      %% &=\frac12 \sum_{i=1}^{k} \log(\sigma^2\lambda_i^{-1} + \eta_i) - k\log \sigma + \frac12\sum_{i=1}^k \log \lambda_i.
    \end{split}
  \end{align}
  Part (3) holds since $\log$ is increasing and $\eta_i \geq 0$.
\end{proof}

\begin{proposition}\label{prop:kkt}
  Let $\tar: \mathbb{R}^m \to \mathbb{R}$, $\psi(\eta) =
  \frac{1}{2}\sum_{i=1}^m \log (1+\sigma^{-2}\lambda_i \eta_i)$, with
  $\lambda_i > 0$ and $\sigma^{2} > 0$. Then the maximum of $\tar$
  subject to $\eta_i \geq 0$ and $\sum\eta_i = m$ is obtained at
  \begin{equation}
  \eta_i = \begin{cases}
    \frac{m}{k} - \sigma^2 \lambda_i^{-1} + \sigma^2 \frac{1}{k} \sum_{j\in A} \lambda_j^{-1} & i \in A \\
    0 & i \in A^c
  \end{cases}
  \end{equation}
  where $A:= \{1\leq i \leq m: \eta_i > 0\}$ and $A^c = \{1,\dots, m\}
    \backslash A$, and $|A|$ is the cardinality of $A$.
\end{proposition}
\begin{proof}
%% \begin{proposition}\label{prop:optimizer}
%%   Let $\{\lambda_i\}_{i=1}^{\infty}$ the sorted eigenvectors of $\fwd
%%   \prcov\fwd^*$, $k\in \mathbb{N}^+$ and
%%   %% Let
%%   %%   \begin{equation}
%%   %%   k = \argmax \left \{ k:\lambda_k^{-1} \leq \sigma^{-2}\frac{m}{k} + \frac{1}{k} \sum_{j=1}^{k}
%%   %%   \lambda_j^{-1} \right \}.
%%   %% \end{equation}

%%   \begin{equation}
%%   \eta_i = \begin{cases}
%%     \frac{m}{k} - \sigma^2 \lambda_i^{-1} + \sigma^2 \frac{1}{k} \sum_{j=1}^k \lambda_j^{-1} & i \leq k \\
%%     0 & i > k 
%%   \end{cases}.
%%   \end{equation}

%%   Let $\obs$ an observation operator such that $\obs^*\obs$ has the
%%   same eigenvectors as $\fwd \prcov \fwd^*$. If $\eta_i > 0$ for
%%   $i\leq k$ and $\obs^*\obs$ has eigenvalues $\eta_i$, then $\obs$
%%   maximizes $\tar$ subject to $\sum_{i=1}^m \eta_i = m$ and to $\rank
%%   \obs = k$.
%% \end{proposition}
  %% \begin{proof}
  %% et $A := \{i: \eta_i > 0\}$. A choice of $A$ where $\eta_i =0,
  %% \eta_j >0$ for some $1 \leq i < j \leq m$ is not optimal (just
  %% switch $\eta_j$ and $\eta_i$ and recall $\{\lambda_i\}$ are arranged
  %% in descending order). Thus, since $\rank \obs = k$ we conclude that,
  %% $\eta_i=0, i>k$.

  Let $\Phi(\eta) = \sum_{i=1}^k \eta_i - m$ and $\Omega_j(\eta) =
   -\eta_j$. Then
  \begin{align*}
    \begin{split}
      \frac{\partial \tar}{\partial \eta_i}  &=
       \frac12 \frac{\sigma^{-2}\lambda_i}{1 + \sigma^{-2} \lambda_i\eta_i} = \frac12 \frac{1}{\sigma^{2}\lambda_i^{-1} + \eta_i} \\
      %
      %
      %
      \frac{\partial\Phi}{\partial \eta_i} &= 1 \\
      %
      %
      %
      \frac{\partial \Omega_j}{\partial \eta_i} &= -\delta_{ij}      
    \end{split}
  \end{align*}

@@

\begin{corollary}\label{cor:descending}
  In Proposition \ref{prop:kkt}, if $\lambda_i$ are arranged in
  descending order, then $A= \{1,\dots,k\}$ for some $1 \leq k \leq
  m$.
\end{corollary}

\begin{corollary}\label{cor:eigenvalues}
  If $\obs$ is optimal and $k := \rank \obs$, then:
  \begin{enumerate}
    \item The covariance of the pushforwad $\fwd_{*} \post$ is $((\fwd
      \prcov \fwd^*)^{-1} + \sigma^{-2} \obs^*\obs)^{-1}$.
    \item The eigenvalues of said operator are
      \begin{equation}
        \eta_i = \begin{cases}
          \frac{m}{k} + \sigma^2 \frac{1}{k} \sum_{j=1}^k \lambda_j^{-1} & i \leq k \\
          \lambda_i^{-1} & i > k 
        \end{cases}.
      \end{equation}
  \end{enumerate}
\end{corollary}

\begin{proof}
  Recall from \eqref{eq:postcov}, that the posterior covariance is
  $\postcov^{-1} = \prcov^{-1} +
  \sigma^{-2}\fwd^*\obs^*\obs\fwd$. Part (1) now follows from Lemma
  \ref{lemma:twice woodbury}. Part (2) follows from Proposition
  \ref{prop:kkt} and Corollary \ref{cor:descending}.
\end{proof}

@@

\main

\begin{proof}
  Part (1) was proved in Corollary \ref{cor:same ev}. Parts (2) and
  (3) were proved in Corollary \ref{cor:true target}. Part (4) is a
  consequence of Corollary \ref{cor:eigenvalues}, with the caveat that
  we did not show that finding $\obs$ so that $\obs^*\obs$ has the
  desired eigenvalues is feasible. To this end, we utilize Lemma
  \ref{lemma:free}: let $M = \diag(\eta_1, \dots, \eta_k)$, diagonal
  with respect to the first $k$ eigenvectors of $\fwd \prcov
  \fwd^*$. We take $\obs := A$ from the Lemma \ref{lemma:free}, and
  this concludes the proof.
\end{proof}

\begin{corollary}\label{cor:equal eigs}
  In a D-optimal design, the posterior covariance of the pushforward
  $\fwd_{*} \post$ is constant for eigenvectors $1 \leq i \leq k$
  where $k =\rank \obs$.
\end{corollary}

@@

    \end{axis}
  \end{tikzpicture}
  \caption{As an illustration of Corollary \ref{cor:equal eigs}, a
    D-optimal design is compared with a sub-optimal design. Both
    designs are allowed three observations. Thus, the blue area has
    accumulated height of $3$ in both plots. A D-optimal design (left)
    increases precision (reduces uncertainty) where it is lowest. A
    sub-optimal design (right) does not. In this example, an
    observation is taken partly in the direction of the third
    eigenvector. The precision for this eigenvector is bigger even
    than the posterior precision of the D-optimal design on the left.}
  \label{fig:optimal vs not}
\end{figure}

Here is a way to understand D-optimal designs with vanishing model
error: Imagine each eigenvector corresponds to a graduated lab
cylinder (Figure \ref{fig:cylinder}). Each cylinder $i$ is filled,
a-priori, with green liquid of $\sigma^2\lambda_i^{-1}$ volume
units. This liquid represnets the prior precision, weighted by
observation noise. The volume of the green liquid is increasing in
$i$. A D-optimal design with $m$ observations ignores cylinders
indexed $m+1$ and above. It has blue liquid of volume $m$ units at its
disposal. It distributes said liquid by repeatedly adding a drop to
whatever cylinder currently has the lowest level of liquid in it
(lowest precision). The result of such a procedure is illustrated in
Figure \ref{fig:optimal vs not}.

@@

\section{Clusterization poses no obstruction to D-optimality}\label{section:clusterization}
Corollary \ref{cor:same ev} tells us that we can realize any design we
want (encoded in $\{\eta_i\}$), given the trace constraint
$\tr{\obs^*\obs} = m$ is satisfied. From Theorem \ref{thm:char} we
know we only need to make the posterior flat where we take
observations (recall figure \ref{fig:optimal vs not}). We are left
with the freedom of assigning any part of $\eta_i$ to any observation
$\meas_j$. Two possible assignments for the same problem are
illustrated in figure \ref{fig:clusterization}. The first exhibits
clusterization, the second does not. It is important to note that
neither arises as a result of some numerical optimization scheme. Both
designs give the same design criterion $\tar$, because their
posteriors are identical. By Theorem \ref{thm:char}, both are
maximizers of the same D-optimal design problem. One exhibits
clusterization, and one does not. Of course, a real-life problem is
more restricted than the model we consider here. Such restriction is
the likely cause of clusterization obseved in, e.g., the toy model of
section \ref{subsec:example}. Exactly how this happens is not
clear. At this point it is the subject of speculation and should be
further studied in the future.

A computer implementation of Lemma \ref{lemma:free} (see code in
Supplementary material) generates the design presented in left panel
Figure \ref{fig:clusterization}, which exhibits clusterization, over
the clusterization-free design presented in the right panel of Figure
\ref{fig:clusterization}. Modifying the numerical values of the prior
does not result in a clusterization-free design. Thus, the preference
to clusterization of Lemma \ref{lemma:free} is robust and
clusterization is a generic property of the model presented in this
study.

@@

\pgfplotstableread{
  Label    prior  $o_1o_2$  $o_3o_4$   $o_5$     topper
  1         0.1       0.95        0.6        1         0.001
  2         0.2       1.0         1.4        0.        0.001
  3         3.5        0          0          0         0.001
}\noclusterization

@@

      \addlegendentry{prior}
      \addlegendentry{$o_1o_2$}
      \addlegendentry{$o_3o_4$}
      \addlegendentry{$o_5$}
      %% \addlegendentry{$o_4$}
      %% \addlegendentry{$o_5$}
      %% \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Clusterization (left) and non-clusterization (right) in
    D-optimal designs. Posterior precision per eigenvector after
    taking observations $\meas_i$ is plotted. Both designs are
    identical for all practical matters --- their posteriors are
    equal. In the left panel we see how an optimal design is achieved
    with repeated observations. $\meas_3 = \meas_5$. This is not
    necessary though, as can be seen in the right panel. Both designs
    achieve the same levels of uncertainty. The left exhibits
    clusterization. The right does not.}
  \label{fig:clusterization}
\end{figure}

>>>>

%% \subsection{The Nonlinear Eigenvalue Problem}\label{subsec:eigenproblem}
The necessary first-order condition for D-optimality
\eqref{eq:conditions}, when $\Sigma= \sigma^2I$ (i.e. when $\modcov =
0$) becomes
\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi,
\end{equation}
with $\Xi$ diagonal. We think of \eqref{eq:eigenproblem} as an
eigenvalue problem for the self-adjoint operator $\sigma^{-2}\fwd
\postcov \fwd^*$, where rows of $\obs$, namely $\meas_j,j=1,\dots, m$,
are eigenvectors. However, $\postcov$ depends on $\obs$ and thus, we
refer to \eqref{eq:eigenproblem} as a \emph{nonlinear} eigenvalue
problem. The following two lemmas allow us to gain better
understanding of the nonlinear eigenvalue problem.

@@

\begin{restatable*}[Simultaneous diagonizability]{lemma}{simdiag}\label{lemma:sim diag}
  Let $\hil$ separable Hilbert space, $C:\hil \to \hil$ self-adjoint
  and $\func_1,\dots,\func_m \in \hil$. Denote $\func^*$ the element
  $\func$ acting as a linear functional. If
  \begin{equation*}
   (C + \sum_{j=1}^m \func_j\func_j^*) \func_l = \xi_l \func_l,\ l = 1,\dots,m
  \end{equation*}
  then $C$ and $\sum_{j=1}^m \func_j \func_j^*$ are simultaneously
  diagonalizable.
\end{restatable*}

\begin{proposition}[Eigenvectors of a D-optimal design]\label{prop:same ev}
  Let $\obs$ satisfy the nonlinear eigenvalue problem
  \eqref{eq:eigenproblem}. Then $\obs^*\obs$ has the same eigenvectors
  as $\fwd \prcov \fwd^*$.
\end{proposition}
\begin{proof}
  \begin{align}\label{eq:mod conditions}
    \begin{split}
      \obs^* \Xi &= \sigma^{-2}\fwd \postcov \fwd^* \obs^*  \text{ (by \eqref{eq:eigenproblem})}\\
      %
      %
      %
      &= \sigma^{-2} \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* \obs^*  \text{ (by \eqref{eq:postcov})} \\
      %
      %
      %
      &= \sigma^{-2} \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1} \obs^* \text{ (by Lemma \ref{lemma:twice woodbury})}.
    \end{split}
  \end{align}

@@

Since we made no assumption regarding the ordering of $\{\lambda_i\}$,
we can denote the corresponding non-zero eigenvalues of $\obs^*\obs$
by $\{\eta_i\}_{i=1}^{k}$ and let $\eta_i = 0$ for $i \geq k+1$.

\begin{proposition}\label{prop:true target}
  Let $\obs$ with $m$ observations satisfy the nonlinear eigenvalue
  problem \eqref{eq:eigenproblem}. Let $\{\eta_i\}_{i=1}^{\infty}$
  eigenvalues of $\obs^*\obs$ and $\{\lambda_i\}_{i=1}^{\infty}$ the
  corresponding eigenvalues of $\fwd \prcov \fwd^*$. Let $k:=\rank
  \obs^*\obs$. Without loss of generality, let $\eta_i > 0$ for $i\leq
  k$ and $\eta_i = 0$ for $i > k$. Then:
  \begin{enumerate}
    \item $k \leq m$ and $\obs^*\obs$ has exactly $k$ positive
      eigenvalues.
    \item
      \begin{equation*}
        \tar(\obs) = \frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i) = \frac12 \sum_{i=1}^{m} \log (1 + \sigma^{-2}\lambda_i\eta_i).
      \end{equation*}
    \item Let $\obs$ a D-optimal design. Then $\eta_i > 0$ for
      eigenvectors corresponding to the $k$ largest $\lambda_i$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  Part (1) is trivial. To see part (2) holds: 
  \begin{align}
    \begin{split}
      \tar(\obs) &= \frac12\log \det \left (I + \sigma^{-2} \prcov^{1/2} \fwd ^* \obs^*
      \obs \fwd \prcov^{1/2}\right )\\% \text{ (by definition)}\\
      %
      &= \frac12 \log \det \left (I + \sigma^{-2} \obs^* \obs \fwd
      \prcov\fwd^* \right ) \text{ (Sylvester's Determinant
      Theorem)}\\
      %
      %
      %
      &=\frac12 \log \prod_{i=1}^{\infty} ( 1 + \sigma^{-2} \lambda_i\eta_i ) \text{ (Proposition \ref{prop:same ev})} \\
      %
      %
      %
      %&=\frac12 \log \left ( \prod_{i=1}^{k} ( \lambda_i^{-1} + \sigma^{-2} \eta_i )\prod_{i=1}^{k} \lambda_i \right )
      %
      %
      %
      &=\frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i). 
      %
      %
      %% &=\frac12 \sum_{i=1}^{k} \log(\sigma^2\lambda_i^{-1} + \eta_i) - k\log \sigma + \frac12\sum_{i=1}^k \log \lambda_i.
    \end{split}
  \end{align}
  Part (3) holds since $\log$ is increasing and $\eta_i \geq 0$.
\end{proof}

\begin{proposition}\label{prop:kkt}
  Let $\tar: \mathbb{R}^m \to \mathbb{R}$, $\psi(\eta) =
  \frac{1}{2}\sum_{i=1}^m \log (1+\sigma^{-2}\lambda_i \eta_i)$, with
  $\lambda_i > 0$ and $\sigma^{2} > 0$. Then the maximum of $\tar$
  subject to $\eta_i \geq 0$ and $\sum\eta_i = m$ is obtained at
  \begin{equation}
  \eta_i = \begin{cases}
    \frac{m}{k} - \sigma^2 \lambda_i^{-1} + \sigma^2 \frac{1}{k} \sum_{j\in A} \lambda_j^{-1} & i \in A \\
    0 & i \in A^c
  \end{cases}
  \end{equation}
  where $A:= \{1\leq i \leq m: \eta_i > 0\}$ and $A^c = \{1,\dots, m\}
    \backslash A$, and $|A|$ is the cardinality of $A$.
\end{proposition}

\begin{proof}
  Let $\Phi(\eta) = \sum_{i=1}^k \eta_i - m$ and $\Omega_j(\eta) =
   -\eta_j$. Then
  \begin{align*}
    \begin{split}
      \frac{\partial \tar}{\partial \eta_i}  &=
       \frac12 \frac{\sigma^{-2}\lambda_i}{1 + \sigma^{-2} \lambda_i\eta_i} = \frac12 \frac{1}{\sigma^{2}\lambda_i^{-1} + \eta_i} \\
      %
      %
      %
      \frac{\partial\Phi}{\partial \eta_i} &= 1 \\
      %
      %
      %
      \frac{\partial \Omega_j}{\partial \eta_i} &= -\delta_{ij}      
    \end{split}
  \end{align*}

@@

@@

\char*

\begin{proof}
  Part (1) is immediate for any observation operator $\obs$ that
  satisfies the unit norm constraint on observation vectors. Part (2)
  was proved in Proposition \ref{prop:same ev}. Part (3) was proved in
  Proposition \ref{prop:true target}.

  Part (4) is a consequence of Propositions \ref{prop:true target} and
  \ref{prop:kkt}, with the caveat that we did not show that finding
  $\obs$ so that $\obs^*\obs$ has the desired eigenvalues is
  feasible. To this end, we utilize Lemma \ref{lemma:free}: let $M =
  \diag(\eta_1, \dots, \eta_k)$, diagonal with respect to the first $k$
  eigenvectors of $\fwd \prcov \fwd^*$. We take $\obs := A$ from the
  Lemma \ref{lemma:free}.

  Recall from \eqref{eq:postcov}, that the posterior precision is
  $\postcov^{-1} = \prcov^{-1} + \sigma^{-2}\fwd^*\obs^*\obs\fwd$. The
  first statement in part (5) now follows from Lemma \ref{lemma:twice
    woodbury}, while the second statement follows from parts (1) and
  (4).
\end{proof}

@@

    \end{axis}
  \end{tikzpicture}
  \caption{A comparison of the eigenvalues of the pushforward
    posterior precision $(\fwd\prcov\fwd^*)^{-1}
    + \sigma^{-2}\obs^*\obs$ for a D-optimal design (left) and a
    sub-optimal design (right). Both designs are allowed three
    observations. We assume $\sigma^2=1$ and thus, the blue area has
    accumulated height of $3$ in both plots. A D-optimal design (left)
    increases precision (reduces uncertainty) where it is lowest. A
    sub-optimal design (right) does not.
    %% In this example, an observation is taken partly in the
    %% direction of the third eigenvector. The precision for this
    %% eigenvector is bigger even than the posterior precision of the
    %% D-optimal design on the left.
  }
  \label{fig:optimal vs not}
\end{figure}

Here is a way to understand D-optimal designs with vanishing model
error: Imagine each eigenvector of $\fwd \prcov \fwd^*$ corresponds to
a graduated lab cylinder (Figure \ref{fig:cylinder}). Each cylinder
$i$ is filled, a-priori, with green liquid of $\lambda_i^{-1}$ volume
units. This liquid represnets the prior precision, and the volume of
the green liquid is increasing in $i$. A D-optimal design with $m$
observations ignores cylinders indexed $m+1$ and above. It has blue
liquid of volume $\sigma^{-2}m$ units at its disposal. It distributes
said liquid by repeatedly adding a drop to whatever cylinder currently
has the lowest level of liquid in it (lowest precision). The result of
such a procedure is illustrated in Figure \ref{fig:optimal vs not}.

@@

\section{Clusterization poses no obstruction to D-optimality}\label{section:clusterization}
We saw that the necessary conditions for a design $\obs$ with $m$
observations to be D-optimal are (1), (2) and (4) of Theorem
\ref{thm:char}. As long as these requirements are satisfied, $\obs$ is
D-optimal. These requirements, however, still leave some freedom in
choosing $\obs$. This freedom may give rise to clusterization, as
illustrated in Figure \ref{fig:clusterization} and explained below.

Consider an inverse problem with $\lambda_1^{-1} = 0.1, \lambda_2^{-1}
= 0.2$ and $\sigma^2=1$. In \eqref{eq:two designs}, we suggest two
D-optimal designs $\obs_1, \obs_2$ such that $\obs_1^*\obs_1 =
\obs_2^*\obs_2$. The former exhibits clusterization ($\meas_3 =
\meas_5 = \ev_1$), while the latter does not. Since both of these
designs are D-optimal, clusterization does not pose any obstruction to
D-optimality.

\begin{equation}\label{eq:two designs}
  \obs_1 =
  \begin{pmatrix}
    \sqrt{0.275}\ev_1 - \sqrt{0.725}\ev_2 \\
    \sqrt{0.275}\ev_1 + \sqrt{0.725}\ev_2 \\
    \ev_1 \\
    \ev_2 \\
    \ev_1 \\    
  \end{pmatrix},
  \obs_2 =
  \begin{pmatrix}
    \sqrt{\frac{3}{8}}\ev_1 - \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{3}{8}}\ev_1 + \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 - \sqrt{\frac{3}{5}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 + \sqrt{\frac{3}{5}}\ev_2 \\
    \ev_1 \\    
  \end{pmatrix}
\end{equation}

A computer implementation of Lemma \ref{lemma:free} (see code in
Supplementary material) generates $\obs_1$, which exhibits
clusterization. Modifying the numerical values of the prior does not
break the clusterization. Thus, the preference to clusterization of
Lemma \ref{lemma:free} is robust. The vast majority of randomized
numerical simulations (see code in Supplementary material) also give
rise to D-optimal designs that exhibit clustrization. We conclude that
clusterization is a generic property of the relaxed model presented in
this study.

%% Of course, a real-life problem is more restricted than the model we
%% consider here. Such restriction is the likely cause of clusterization
%% obseved in, e.g., the toy model of section
%% \ref{subsec:example}. Exactly how this happens is not clear. At this
%% point it is the subject of speculation and should be further studied
%% in the future.

@@

\pgfplotstableread{
  Label    prior  $o_1o_2$  $o_3o_4$   $o_5$     topper
  1         0.1    0.75     0.8        1         0.001
  2         0.2    1.25     1.2        0         0.001
  3         3.5    0        0          0         0.001
}\noclusterization

@@

      \addlegendentry{prior}
      \addlegendentry{$o_1o_2$}
      \addlegendentry{$o_3o_4$}
      \addlegendentry{$o_5$}
      %% \addlegendentry{$o_4$}
      %% \addlegendentry{$o_5$}
      %% \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Clusterization (left, $\obs_1$ in \eqref{eq:two designs})
    and non-clusterization (right, $\obs_2$ in \eqref{eq:two designs})
    in D-optimal designs. Posterior precision per eigenvector is
    plotted for $\obs_1$ and $\obs_2$ from \eqref{eq:two
      designs}. Both designs are identical for all practical matters
    --- their posteriors are equal. The left panel shows a D-optimal
    design with clusterization: $\meas_3 = \meas_5$. The right panel
    shows a D-optimal design without clusterization. The contribution
    to the posterior precision due to $\meas_1$ and $\meas_2$ is
    denoted $\meas_1\meas_2$. These contributions are plotted together
    because $\meas_1$ and $\meas_2$ are not in the direction of any
    eigenvector. Similarly for $\meas_3\meas_4$.}
  \label{fig:clusterization}
\end{figure}
